Lossy compression of DNA sequencing quality data

### 2015-05-15 manuscript/q-compression.tex
revised a sentence in the discussion
refinement of the discussion section

We have examined several simple and general approaches for lossy
compression of read quality values, and their effect on bioinformatic
analyses. Our results demonstrate that some compression methods can
result in transformed quality values that are quite useful, and in
some cases advantageous, compared to original uncompressed values.

Downstream applications dictate the relative performance lossy
compression methods.  Many bioinformatics tools proved to be robust to
moderate information loss in base-call quality. The use of compression
transformed quality values in quality control demonstrate that several
compression methods result in approximately the same total number
base-calls passing through the combined filtering and trimming
processes as when using original quality values, though the absolute
and relative number of reads filtered and base-calls trimmed may
differ. Genome assembly appears robust to standard sequencing errors,
and compression transformed quality values, particularly those
associated with profiling-based compression, result in assemblies
better than those based on original uncompressed quality values by
most measures.

@@

Quality values for base-calls in \textsc{dna} sequencing are not
sacrosanct; they are representations of estimates for measurement
(detection) error determined by various mechanisms including signal
processing characteristics and algorithms. The true worth of quality
values lies in the utility they provide in bioinformatic analyses. In
some cases, as demonstrated here, this value is enhanced through the
quality values transformed through compression processes. Thus
appropriate compression of quality values can both reduce associated
memory cost and improve analysis results.

>>>>

We have examined several simple and general approaches for lossy
compression of read quality values, and their effect on bioinformatic
analyses. Our results demonstrate that some compression methods can
result in quality values that are quite useful, and in some cases
advantageous, compared to original uncompressed values.

Downstream applications dictate the relative performance lossy
compression methods.  Some bioinformatics tools proved to be robust to
moderate information loss in base-call quality. The use of quality
values modified via several compression/decompression processes result
in approximately the same total number base-calls passing through the
combined filtering and trimming processes as when using original
quality values, though the absolute and relative number of reads
filtered and base-calls trimmed may differ. Genome assembly appears
robust to standard sequencing errors, and compression transformed
quality values, particularly those associated with profiling-based
compression, result in assemblies better than those based on original
uncompressed quality values by most measures.

@@

Quality values for base-calls in \textsc{dna} sequencing are not
sacrosanct; they are representations of estimates for measurement
(detection) error determined by various mechanisms including signal
processing characteristics and algorithms. The true worth of quality
values lies in the utility they provide in bioinformatic analyses. In
some cases, as demonstrated here, this value is enhanced through
transformation associated with compression. Thus appropriate
compression of quality values can both reduce associated memory cost
and improve analysis results.
